---
title: "Final Project Report"
author: "Gajin Kim & Jayant Shah"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(RColorBrewer)
```

```{}
Project background:
Our group will perform a high-level analysis of the effects of Covid-19 on average temperatures in different cities across the United States before, during, and after the pandemic. The two cities we will be focusing on are New York City, NY and Waterloo, IA. We determined that incorporating city datasets of two vasty different population densities would enable us to better understand any underlying trends.
```
```{}
Questions raised:
1. How has the average temperature changed between the start and end of Covid-19?
2. How has the temperature of densely populated cities (i.e. NYC, New York) been affected by Covid-19?
3. How has the temperature of a rural city (i.e. Waterloo, Iowa) been affected by Covid-19?
4. Main Question: How effective has the quarantine been in reducing the daily mean temperature?
```
```{}
Obtaining the Dataset:
We obtained our raw datasets through an online repository containing different features for various airports across the United States which can be accessed here: https://www.ncei.noaa.gov/access/search/data-search/. We utilized the Waterloo Municipal Airport and JFK International Airport as our Waterloo and NYC datasets, respectively. Additionally, each data frame represented a year for that given airport (e.g. Waterloo_Muncipial 2019). Utilizing this dataset was quite convenient as it tracked numerous useful features that capture different temperatures and weather patterns across different years, however many column values were either undefined or cryptic.

Cleaning the Dataset:
Our datasets consisted of over 100 columns and 50,000 rows across the following 6 dataframes: 
- NYC 2019
- NYC 2020
- NYC 2021
- Waterloo 2019
- Waterloo 2020
- Waterloo 2021

Each row initially represented a single hour of a given day across 3 different years, doubled (as we had 2 cities). We first combined each of the dataset by city, giving us 2 data frames:
- NYC Combined
- Waterloo Combined

From here, we had to address the issue that each row representing a single hour. To fix this issue, we created a new date column and averaged values by day and collapsed our rows. Basically, each row now represents a single day rather than one hour. Finally, we removed columns that were either undefined, contained indecipherable values (due to a lack of proper documentation), or were simply unnecessary for analysis. We also wrote our cleaned datasets to a new csv so we didn't have to reclean them every time we wanted to perform analysis.
```
```{r}
# Load our cleaned data sets
waterloo_clean <- read.csv('./datasets/Waterloo_Municipal_IowaClean.csv')
nyc_clean <- read.csv('./datasets/JFK_International_NewYorkClean.csv')
```

***

**Question 1.**

We have two data sets for two different cities: Waterloo and New York.

To compare mean values of temperatures for pre-, during-, and post-quarantine periods, the data has loosely split up by year.

- Pre-Quarantine Period is year 2019
- During-Quarantine Period is year 2020
- Post-Quarantine Period is year 2021
```{r echo=FALSE}
data_w<-split(waterloo_clean, waterloo_clean$Year)

waterloo_pre<-data_w$`2019`
waterloo_during<-data_w$`2020`
waterloo_post<-data_w$`2021`

w_pre_monthly<-waterloo_pre[!is.na(waterloo_pre$MonthlyMeanTemperature),]
w_d_monthly<-waterloo_during[!is.na(waterloo_during$MonthlyMeanTemperature),]
w_post_monthly<-waterloo_post[!is.na(waterloo_post$MonthlyMeanTemperature),]

# For New York:
data_nyc<-split(nyc_clean, nyc_clean$Year)

nyc_pre<-data_nyc$`2019`
nyc_during<-data_nyc$`2020`
nyc_post<-data_nyc$`2021`

nyc_pre_monthly<- nyc_pre[!is.na(nyc_pre$MonthlyMeanTemperature),]
nyc_d_monthly<- nyc_during[!is.na(nyc_during$MonthlyMeanTemperature),]
nyc_post_monthly<- nyc_post[!is.na(nyc_post$MonthlyMeanTemperature),]
```
```{r echo=FALSE}
w_summary<-round(cbind(summary(w_pre_monthly$MonthlyMeanTemperature), summary(w_d_monthly$MonthlyMeanTemperature), summary(w_post_monthly$MonthlyMeanTemperature)), 2)
colnames(w_summary)=c("2019", "2020", "2021")
```

In the summary below, we can observe the average temperature in Waterloo by month between 2019 and 2021. 

```{r}
w_summary
```
```{r echo=FALSE}
# For New York:
nyc_summary<-round(cbind(summary(nyc_pre_monthly$MonthlyMeanTemperature), summary(nyc_d_monthly$MonthlyMeanTemperature), summary(nyc_post_monthly$MonthlyMeanTemperature)), 2)
colnames(nyc_summary)=c("2019", "2020", "2021")
```

Likewise, we can observe the average temperature in New York City by month between 2019 and 2021 in the figure below. 

```{r}
nyc_summary
```

We will plot the corresponding graph to compare the Monthly Average Temperature for both cities in order to observe any significant differences between our graphs.

```{r echo=FALSE}
# Waterloo
plot(w_pre_monthly$Month, w_pre_monthly$MonthlyMeanTemperature, type = "b", col= "red", xlab = "Month", ylab = "Average Temperature", main = "Average Monthly Temperature in Waterloo in years 2019 - 2021")

lines(w_d_monthly$Month, w_d_monthly$MonthlyMeanTemperature, type = "b", col = "blue")
lines(w_post_monthly$Month, w_post_monthly$MonthlyMeanTemperature, type = "b", col = "green")
legend("topleft", legend = c(2019, 2020, 2021), col = c("red", "blue", "green"), lty = 1:2, cex = 0.8)

# New York City
plot(nyc_pre_monthly$Month, nyc_pre_monthly$MonthlyMeanTemperature, type = "b", col= "red", xlab = "Month", ylab = "Average Temperature", main = "Average Monthly Temperature in New York in years 2019 - 2021")

lines(nyc_d_monthly$Month, nyc_d_monthly$MonthlyMeanTemperature, type = "b", col = "blue")
lines(nyc_post_monthly$Month, nyc_post_monthly$MonthlyMeanTemperature, type = "b", col = "green")
legend("topleft", legend = c(2019, 2020, 2021), col = c("red", "blue", "green"), lty = 1:2, cex = 0.8)
```

As we can see, we cannot say if there is a difference between the average monthly temperature in pre-, during-, and post-covid periods in Waterloo or New York. Thus, we heed to test hypothesis about the equivalence of mean values for these periods. We will use t-test:

```{r echo=FALSE}
t_w1<-t.test(w_pre_monthly$MonthlyMeanTemperature,w_d_monthly$MonthlyMeanTemperature)
t_w2<-t.test(w_pre_monthly$MonthlyMeanTemperature,w_post_monthly$MonthlyMeanTemperature)
t_w3<-t.test(w_d_monthly$MonthlyMeanTemperature,w_post_monthly$MonthlyMeanTemperature)
```
```{r}
t_w1
```
As we can see, the p-value is 0.7754, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for pre-Covid and during-Covid period in Waterloo is equal to zero.
```{r}
t_w2
```
As we can see, the p-value is 0.541, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for pre-Covid and post-Covid period in Waterloo is equal to zero.
```{r}
t_w3
```
As we can see, the p-value is 0.7059, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for during-Covid and post-Covid period in Waterloo is equal to zero.

```{r echo=FALSE}
#For New York, we have:
t_ny1<-t.test(nyc_pre_monthly$MonthlyMeanTemperature,nyc_d_monthly$MonthlyMeanTemperature)
t_ny2<-t.test(nyc_pre_monthly$MonthlyMeanTemperature,nyc_post_monthly$MonthlyMeanTemperature)
t_ny3<-t.test(nyc_d_monthly$MonthlyMeanTemperature,nyc_post_monthly$MonthlyMeanTemperature)
```
```{r}
t_ny1
```
As we can see, the p-value is 0.8145, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for pre-Covid and during-Covid period in New York is equal to zero.

```{r}
t_ny2
```
As we can see, the p-value is 0.6132, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for pre-Covid and post-Covid period in New York is equal to zero.

```{r}
t_ny3
```
As we can see, the p-value is 0.7576, which is greater than 0.05. It means that we cannot reject the null-hypothesis about equality of means. Thus, the true difference in means for during-Covid and post-Covid period in New York is equal to zero.

Thus, as we can see, the average monthly temperature in pre-, during-, and post-COVID periods does not have statistically significant difference, i.e. we may conclude that the average monthly temperature did not change significantly in COVID period both in Waterloo and New York.

***

**Question 2.**

We will being by utilizing the 'Year' variable as a factor.
```{r}
nyc_clean$Year<-as.factor(nyc_clean$Year)
```
By using this data and 'Year' as factor, we can build a boxplot to observe the average daily temperature in New York in years 2019 (pre-COVID time), 2020 (during-COVID time), and 2021 (post-COVID time):

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = nyc_clean, aes(y = DailyAverageDryBulbTemperature, x = Year), fill = "class") + 
geom_boxplot()+
xlab("Year")+
ylab("Daily Average Temperature")+
ggtitle("Daily Average Temperature in New York by Year (2019 - 2021)")
```
As we can see from the boxplot, we can assume that the average daily temperature in New York during COVID-19 has changed: it is slightly greater for year 2020 compared to 2019; it's greater for year 2021 compared to 2019 and 2020.
The summary statistics:

```{r echo=FALSE}
tapply(nyc_clean$DailyAverageDryBulbTemperature, nyc_clean$Year, function(x) format(summary(x)))
```
As we can see from the summary statistic, our assumption should be right. Let check it with t-test:

```{r echo=FALSE}
tt_ny1<-t.test(nyc_pre$DailyAverageDryBulbTemperature, nyc_during$DailyAverageDryBulbTemperature)
tt_ny2<-t.test(nyc_pre$DailyAverageDryBulbTemperature, nyc_post$DailyAverageDryBulbTemperature)
tt_ny3<-t.test(nyc_during$DailyAverageDryBulbTemperature, nyc_post$DailyAverageDryBulbTemperature)
```

The results are:
```{r}
tt_ny1
```

As we can see, the p-value of the test is 0.2219, which is greater than 0.05. It means that at 95% confidence level, we cannot reject the null-hypothesis: the true difference in means is equal to zero. Thus, there is no difference in daily average temperature in New York in pre- and during-COVID periods.

```{r}
tt_ny2
```

As we can see, the p-vlue is 0.01615, which is lower than 0.05. Thus, we should reject the null-hypothesis about equality of means. Thus, there is a statistically significant difference in pre- and post- daily average temperature in New York. We can check if our assumption is true and the post-COVID average daily temperature is higher than in pre-COVID period in NY:

```{r}
tt_ny2_new<-t.test(nyc_post$DailyAverageDryBulbTemperature, nyc_pre$DailyAverageDryBulbTemperature, alternative = "greater")
tt_ny2_new
```

As we can see, the p-value is 0.008075, which is lower than 0.05: we should reject the null-hypothesis. It means that the true difference in means between post-COVID daily average temperature and pre-COVID daily average temperature is greater than zero. Thus, the post-COVID daily average temperature is higher than the pre-COVID daily average temperature in New York.

Let see what we have for during- and post-COVID daily average temperature difference:

```{r}
tt_ny3
```

As we can see, the p-value is 0.1869, which is greater than 0.05, which means that we cannot reject the null-hypothesis. Thus, the difference between the mean values for daily average temperature in during- and post-COVID periods are equal to zero. Thus, there is no statistically significant difference in daily average temperatures in during- and post-COVID periods.

Thus, we can conclude that the post-COVID daily average temperature in New York is higher than in pre-COVID period, but it's not different from  during-COVID period. The daily average temperature in pre-COVID and during-COVID period are not significantly different.

***

**Question 3.**

```{r}
waterloo_clean$Year<-as.factor(waterloo_clean$Year)
```

By using this data and 'Year' as factor, we can build a boxplot to see the average daily temperature in Waterloo in years 2019 (pre-COVID time), 2020 (during-COVID time), and 2021 (post-COVID time):

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = waterloo_clean, aes(y = DailyAverageDryBulbTemperature, x = Year), fill = "class") + 
geom_boxplot()+
xlab("Year")+
ylab("Daily Average Temperature")+
ggtitle("Daily Average Temperature in Waterloo by Year (2019 - 2021)")

```

As we can see from the boxplot, we can assume that the average daily temperature in this city during COVID-19 has changed: it is slightly greater for year 2020 compared to 2019; it's greater for year 2021 compared to 2019 and slightly greater compared to year 2020.
The summary statistics:

```{r}
tapply(waterloo_clean$DailyAverageDryBulbTemperature, waterloo_clean$Year, function(x) format(summary(x)))
```
As we can see from the summary statistic, our assumption should be right. Let check it with t-test:

```{r echo=FALSE}
tt_w1<-t.test(waterloo_pre$DailyAverageDryBulbTemperature, waterloo_during$DailyAverageDryBulbTemperature)
tt_w2<-t.test(waterloo_pre$DailyAverageDryBulbTemperature, waterloo_post$DailyAverageDryBulbTemperature)
tt_w3<-t.test(waterloo_during$DailyAverageDryBulbTemperature, waterloo_post$DailyAverageDryBulbTemperature)
```

The results are:
```{r}
tt_w1
```

As we can see, the p-value of the test is 0.1414, which is greater than 0.05. It means that at 95% confidence level, we cannot reject the null-hypothesis: the true difference in means is equal to zero. Thus, there is no difference in daily average temperature in Waterloo in pre- and during-COVID periods.

```{r}
tt_w2
```

As we can see, the p-value is 0.002768, which is lower than 0.05. Thus, we should reject the null-hypothesis about equality of means. Thus, there is a statistically significant difference in pre- and post- daily average temperature in Waterloo. We can check if our assumption is true and the post-COVID average daily temperature is higher than in pre-COVID period in Waterloo:

```{r}
tt_w2_new<-t.test(waterloo_post$DailyAverageDryBulbTemperature, waterloo_pre$DailyAverageDryBulbTemperature, alternative = "greater")
tt_w2_new
```

As we can see, the p-value is 0.001384, which is lower than 0.05: we should reject the null-hypothesis. It means that the true difference in means between post-COVID daily average temperature and pre-COVID daily average temperature is greater than zero. Thus, the post-COVID daily average temperature is higher than the pre-COVID daily average temperature in Waterloo.

Let see what we have for during- and post-COVID daily average temperature difference:
```{r}
tt_w3
```

As we can see, the p-value is 0.08595, which is greater than 0.05. But it is lower than 0.1, which means that at 90% of confidence we can say that the mean values of daily average temperature for during- and post-COVID periods are different. We can check this:

```{r}
tt_w3_new<-t.test(waterloo_post$DailyAverageDryBulbTemperature, waterloo_during$DailyAverageDryBulbTemperature, alternative = "greater")
tt_w3_new
```

As we can see, the p-value is 0.04297, which is lower than 0.05: we should reject the null-hypothesis. It means that the true difference in means between mean values of daily average temperature in post- and during-COVID period is greater than zero. Thus, the mean value of post-COVID daily average temperature is higher than the mean value of during-COVID daily average temperature in Waterloo.

Thus, we can conclude that the post-COVID daily average temperature in Waterloo is higher than in pre-COVID and during-COVID periods. The daily average temperature in pre-COVID and during-COVID period is not significantly different.

***

**Question 4. & Conclusion to Project Analysis**

Based on the previous analysis, the quarantine has not been effective in reducing daily mean temperature or average temperature in general. Post-COVID average daily temperature is higher than pre-COVID average daily temperature both for New York and Waterloo. We suspect that this result can be attributed to the fact that despite people were staying indoors more often during/post Covid-19, they were also utilizing electricity  and ordering food more frequently than before.
```{}
Contribution Statements:

Gajin Kim -
Both Jayant and I worked on all parts of the project from beginning, middle, and finish. That said, we determined that it would be more optimal for each of us to prioritize certains parts of the project given our schedules. Some of the things I primarily focused on were procuring adequate datasets, cleaning the datasets, and creating the presentation. We both made contributions to the project proposal and analysis.
```